---
title: "Ciprofloxacin resistance in Klebsiella pneumoniae"
author: "Iana Amke"
date: "Last compiled on `r format(Sys.time(), '%d/%m/%y')`"
output: 
  html_document: 
    df_print: paged
    fig_width: 9
    fig_height: 8
    mode: selfcontained
    highlight: pygments
    theme: 
      bootswatch: flatly
      code_font:
        google: "JetBrains Mono"
editor_options: 
  chunk_output_type: inline
---
<style type="text/css">
.main-container {
  max-width: 1600px !important;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = 'png', echo = TRUE, message = FALSE, warning = FALSE,
                      cache = FALSE, fig.align = 'center')
```

```{r functions, include=FALSE}
# functions 
# 1. upset_violin_plot to create violin + UpSet plots for AST measurements and fluoroquinolone resistance mechanisms

# create violin + UpSet plots for AST measurements and fluoroquinolone resistance mechanisms
upset_violin_plot <- function(df, x, y, z) {
# df - dataframe; x - Flq resistance mechanism column; y - Laboratory typing method column
# z - Testing.standard
  require(dplyr, ggplot2, ggupset)
  
  # checks if lab typing method is MIC or disk diffusion 
  if (grepl('MIC.*$', df[y])) {
    # checks if testing standard is EUCAST or CLSI
    if (grepl('EUCAST', df[z])) {
      df %>%
        group_by(strain, Measurement, Resistance.phenotype) %>% 
        # !! sym() allows one to pass a column name as an argument
        summarize(res_mech_list = list(!! sym(x))) %>%
        ggplot(aes(res_mech_list, Measurement)) + geom_violin() + geom_count(aes(colour = Resistance.phenotype)) + 
        # EUCAST cut-off of 1.0 mg/L is resistant
        geom_hline(aes(yintercept = 1.0, linetype = "resistant"), alpha = 0.6, color = "black") +
        # EUCAST cut-off of 0.25 mg/L is susceptible
        geom_hline(aes(yintercept = 0.25, linetype = "susceptible"), alpha = 0.6, color = "black") +
        labs(x = "Fluoroquinolone resistance determinants", y = "Ciprofloxacin MIC measurement (mg/L)", 
             size = "Number of strains", linetype = "MIC breakpoints", colour = "Phenotype") +
        # convert y-axis to log2 scale and labels to 3 dp
        scale_y_continuous(trans = log2_trans(), breaks = 2^(-5:9), labels = function(x) round(as.numeric(x), digits = 3)) +
        scale_linetype_manual(values = c("solid", "dotted")) +
        scale_x_upset() +
        theme_combmatrix(combmatrix.panel.line.size = 0.8)
    } else {
      df %>%
        group_by(strain, Measurement, Resistance.phenotype) %>% 
        # !! sym() allows one to pass a column name as an argument
        summarize(res_mech_list = list(!! sym(x))) %>%
        ggplot(aes(res_mech_list, Measurement)) + geom_violin() + geom_count(aes(colour = Resistance.phenotype)) + 
        # CLSI cut-off of 1.0 mg/L is resistant
        geom_hline(aes(yintercept = 1.0, linetype = "resistant"), alpha = 0.6, color = "black") +
        # CLSI cut-off of 0.25 mg/L is susceptible
        geom_hline(aes(yintercept = 0.25, linetype = "susceptible"), alpha = 0.6, color = "black") +
        labs(x = "Fluoroquinolone resistance determinants", y = "Ciprofloxacin MIC measurement (mg/L)", 
             size = "Number of strains", linetype = "MIC breakpoints", colour = "Phenotype") +
        # convert y-axis to log2 scale and labels to 3 dp
        scale_y_continuous(trans = log2_trans(), breaks = 2^(-5:9), labels = function(x) round(as.numeric(x), digits = 3)) +
        scale_linetype_manual(values = c("solid", "dotted")) +
        scale_x_upset() +
        theme_combmatrix(combmatrix.panel.line.size = 0.8)
    }
  } else {
    if (grepl('EUCAST', df[z])) {
      df %>%
        group_by(strain, Measurement, Resistance.phenotype) %>% 
        summarize(res_mech_list = list(!! sym(x))) %>%
        ggplot(aes(res_mech_list, Measurement)) + geom_violin() + geom_count(aes(colour = Resistance.phenotype)) + 
        # EUCAST cut-off of 21mm is resistant
        geom_hline(aes(yintercept = 21, linetype = "resistant"), alpha = 0.6, color = "black") +
        # EUCAST cut-off of 25mm is susceptible
        geom_hline(aes(yintercept = 25, linetype = "susceptible"), color = "black") +
        labs(title = "Fluoroquinolone resistance determinant combinations\nat various disk diffusion measurements", x = 
               "Fluoroquinolone resistance determinants", y = "Ciprofloxacin disk diffusion measurement (mm)", 
             size = "Number of strains", linetype = "Disk diffusion breakpoints", colour = "Phenotype") +
        scale_linetype_manual(values = c("solid", "dotted")) +
        scale_x_upset() +
        theme_combmatrix(combmatrix.panel.line.size = 0.8)
    } else {
      df %>%
        group_by(strain, Measurement, Resistance.phenotype) %>% 
        summarize(res_mech_list = list(!! sym(x))) %>%
        ggplot(aes(res_mech_list, Measurement)) + geom_violin() + geom_count(aes(colour = Resistance.phenotype)) + 
        # CLSI cut-off of 21mm is resistant
        geom_hline(aes(yintercept = 21, linetype = "resistant"), alpha = 0.6, color = "black") +
        # CLSI cut-off of 26mm is susceptible
        geom_hline(aes(yintercept = 26, linetype = "susceptible"), color = "black") +
        labs(title = "Fluoroquinolone resistance determinant combinations\nat various disk diffusion measurements", x = 
               "Fluoroquinolone resistance determinants", y = "Ciprofloxacin disk diffusion measurement (mm)", 
             size = "Number of strains", linetype = "Disk diffusion breakpoints", colour = "Phenotype") +
        scale_linetype_manual(values = c("solid", "dotted")) +
        scale_x_upset() +
        theme_combmatrix(combmatrix.panel.line.size = 0.8)
    }
  }
}

# 2. upset_bar_plot to create percent stacked bar charts + UpSet plots for AST measurements and fluoroquinolone resistance mechanisms

# create percent stacked barchart + UpSet plots for AST measurements and fluoroquinolone resistance mechanisms
upset_bar_plot <- function(df, x) {
# df - dataframe; x - Flq resistance mechanism column
  
  require(dplyr, ggplot2, ggupset)
  
  df %>%
    group_by(strain, Measurement) %>% 
    # !! sym() allows one to pass a column name as an argument
    summarize(res_mech_list = list(!! sym(x)), Resistance.phenotype) %>%
    ggplot(aes(res_mech_list, Measurement, fill = Resistance.phenotype)) + geom_bar(position="fill", stat="identity") +
    labs(x = "Fluoroquinolone resistance determinants", y = "% phenotype",
         fill = "Phenotype") +
    scale_x_upset() +
    theme_combmatrix(combmatrix.panel.line.size = 0.8)
}

# 3. group_res_mech to group resistance determinants

# function to group some resistance determinants
group_res_mech <- function(df, x) {
# df - data frame; x - column(s) with resistance determinanats
# {{ }} and := allows one to pass a column name as a function parameter
  
  require(dplyr)
  
  rare_res_qnr_genes <- c("qnrB7", "qnrE2", "qnrS2", "qnrVC1", "qnrVC6")

  df %>%
    # based on statistical analysis (chunk proportion), combine QRDR mutations to have just 4 types (GyrA-83, GyrA-87, ParC-80, ParC-84)
    # except GyrA-87H which is found in one strain and is sensitive
    mutate({{ x }} := replace({{ x }}, startsWith({{ x }}, "GyrA-83"), "GyrA-83")) %>%
    mutate({{ x }} := case_when(startsWith({{ x }}, "GyrA-87") & {{ x }} != "GyrA-87H" ~ "GyrA-87",
                                 TRUE ~ {{ x }})) %>%
    mutate({{ x }} := replace({{ x }}, startsWith({{ x }}, "ParC-80"), "ParC-80")) %>%
    mutate({{ x }} := replace({{ x }}, startsWith({{ x }}, "ParC-84"), "ParC-84")) %>%
    # also combine the following acquired genes
    mutate({{ x }} := replace({{ x }}, startsWith({{ x }}, "qepA2"), "qepA2")) %>%
    mutate({{ x }} := replace({{ x }}, startsWith({{ x }}, "qnrA1"), "qnrA1")) %>%
    mutate({{ x }} := replace({{ x }}, startsWith({{ x }}, "qnrB1"), "qnrB1")) %>%
    mutate({{ x }} := replace({{ x }}, startsWith({{ x }}, "qnrB19"), "qnrB19")) %>%
    mutate({{ x }} := replace({{ x }}, startsWith({{ x }}, "qnrB2"), "qnrB2")) %>%
    mutate({{ x }} := replace({{ x }}, startsWith({{ x }}, "qnrS1"), "qnrS1")) %>%
    mutate({{ x }} := replace({{ x }}, grepl(paste(rare_res_qnr_genes, collapse = "|"), {{ x }}), "rare_res_qnr"))
}

```

# Preliminary analysis of sample data
The data is made up of assembled *K. pneumoniae* genomes with matched antibiogram data. The first step of data analysis involved filtering for ciprofloxacin and mechanisms that contribute to fluoroquinolone resistance i.e. *gyrA* and *parC* mutations found in the Flq_mutations column.  

Afterwards, the data was visualised using bar plots which showed sample count per dataset and sample count for each of the laboratory typing methods used i.e MIC broth and agar dilution and disk diffusion and violin plots in combination with UpSet plots which showed the distribution of fluoroquinolone mutations in the genomes being analysed. Disk diffusion and MIC measurements were also checked to make sure they were valid - those not valid were filtered out.

## Load required libraries
```{r}
library(here)
library(tidyverse)
library(scales)
library(ggupset)
library(RColorBrewer)
library(collapse)
library(DT)
library(naniar)
library(data.table)
library(rebus)
```

## Load antibiogram data and filter for ciprofloxacin
```{r}
# read antibiogram data
antibiogram <- read.delim(file = "antibiogram_combined.tsv", 
                                header = TRUE, sep = "\t")

# filter for ciprofloxacin and select Flq_mutations column
cipro_antibiogram <- antibiogram %>%
  filter(Antibiotic == "ciprofloxacin") %>%
  select(strain:ST, Flq_mutations, Antibiotic:Measurement.Round) %>%
  mutate(Resistance.phenotype = case_when(grepl('MIC.*$', Laboratory.Typing.Method) & Measurement <= 0.25 ~ 
                                            "susceptible",
                                          grepl('MIC.*$', Laboratory.Typing.Method) & Testing.standard == 
                                            "EUCAST" & Measurement > 0.5 ~ "resistant",
                                          grepl('MIC.*$', Laboratory.Typing.Method) & Testing.standard == 
                                            "CLSI" & Measurement >= 1 ~ "resistant",
                                          Laboratory.Typing.Method == "Disk diffusion" & Testing.standard == 
                                            "EUCAST" & Measurement >= 25 ~ "susceptible",
                                          Laboratory.Typing.Method == "Disk diffusion" & Testing.standard == 
                                            "EUCAST" & Measurement < 22 ~ "resistant",
                                          Laboratory.Typing.Method == "Disk diffusion" & Testing.standard == 
                                            "CLSI" & Measurement >= 26 ~ "susceptible",
                                          Laboratory.Typing.Method == "Disk diffusion" & Testing.standard == 
                                            "CLSI" & Measurement <= 21 ~ "resistant",
                                          TRUE ~ "intermediate")) %>%
  # filter out strains with measurement of 0.12 and 0.75 (no such MIC measurements)
  filter(Measurement != 0.12 & Measurement != 0.75)

#write_delim(cipro_antibiogram, file = "cipro_antibiogram.tsv", delim = "\t")
```

## Summaries {.tabset .tabset-pills}
### Sample count per dataset
```{r}
cipro_antibiogram %>%
  group_by(index) %>%
  ggplot(aes(x = index)) +
  geom_bar(fill = "#ff8000", alpha = 0.6) +
  geom_text(aes(label = ..count..), stat = "count", nudge_y = 60) +
  labs(title = "Total sample count per dataset", y = "Total sample count", fill = "Lab typing method") +
  theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

### Total sample counts vs measurements for different lab typing methods {.tabset .tabset-pills}
#### MIC (broth dilution) {.tabset .tabset-pills}
##### CLSI
```{r}
cipro_antibiogram %>%
  group_by(Laboratory.Typing.Method) %>%
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "CLSI") %>% 
  ggplot(aes(x = factor(Measurement), fill = index)) + geom_bar() + 
  # CLSI cut-off of 1.0 mg/L is resistant
  geom_vline(aes(xintercept = 5, linetype = "resistant"), alpha = 0.6, color = "black") +
  # CLSI cut-off of 0.25 mg/L is susceptible
  geom_vline(aes(xintercept = 3, linetype = "susceptible"), color = "black") +
  labs(title = "Total sample count at various MIC (broth dilution) measurements", x = "Measurement (mg/L)", y = "Frequency", fill = "Dataset", 
       linetype = "MIC breakpoints") +
  scale_linetype_manual(values = c("solid", "dotted")) 
```

##### EUCAST
```{r}
cipro_antibiogram %>%
  group_by(Laboratory.Typing.Method) %>%
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "EUCAST") %>% 
  ggplot(aes(x = factor(Measurement), fill = index)) + geom_bar() + 
  # EUCAST cut-off of 1.0 mg/L is resistant
  geom_vline(aes(xintercept = 8, linetype = "resistant"), alpha = 0.6, color = "black") +
  # EUCAST cut-off of 0.25 mg/L is susceptible
  geom_vline(aes(xintercept = 5, linetype = "susceptible"), color = "black") +
  labs(title = "Total sample count at various MIC (broth dilution) measurements", x = "Measurement (mg/L)", y = "Frequency", fill = "Dataset", 
       linetype = "MIC breakpoints") +
  scale_linetype_manual(values = c("solid", "dotted")) 
```

#### MIC (agar dilution)
##### EUCAST
```{r}
# only EUCAST used as testing standard
cipro_antibiogram %>%
  group_by(Laboratory.Typing.Method) %>%
  filter(Laboratory.Typing.Method == "MIC (agar dilution)") %>% 
  ggplot(aes(x = factor(Measurement), fill = index)) + geom_bar() + 
  # EUCAST cut-off of 1.0 mg/L is resistant
  geom_vline(aes(xintercept = 6, linetype = "resistant"), alpha = 0.6, color = "black") +
  # EUCAST cut-off of 0.25 mg/L is susceptible
  geom_vline(aes(xintercept = 4, linetype = "susceptible"), color = "black") +
  labs(title = "Total sample count at various MIC (agar dilution) measurements", x = "Measurement (mg/L)", y = "Frequency", fill = "Dataset", 
       linetype = "MIC breakpoints") +
  # round off x-axis labels to 2dp
  scale_x_discrete(labels = function(x) round(as.numeric(x), digits = 2)) +
  scale_linetype_manual(values = c("solid", "dotted")) 
```

#### Disk diffusion {.tabset .tabset-pills}
##### CLSI
```{r}
cipro_antibiogram %>%
  group_by(Laboratory.Typing.Method) %>%
  filter(Laboratory.Typing.Method == "Disk diffusion") %>%
  filter(Testing.standard == "CLSI") %>%
  ggplot(aes(Measurement, fill = index)) + geom_bar() + 
  # CLSI cut-off of 21mm is resistant
  geom_vline(aes(xintercept = 21, linetype = "resistant"), color = "black") +
  # CLSI cut-off of 26mm is susceptible
  geom_vline(aes(xintercept = 26, linetype = "susceptible"), color = "black") +
  labs(title = "Total sample count at various disk diffusion measurements", x = "Measurement (mm)", y = "Frequency", fill = "Dataset", 
       linetype = "Disk diffusion \nbreakpoints") +
  scale_linetype_manual(values = c("solid", "dotted"))
```

##### EUCAST
```{r}
cipro_antibiogram %>%
  group_by(Laboratory.Typing.Method) %>%
  filter(Laboratory.Typing.Method == "Disk diffusion") %>%
  filter(Testing.standard == "EUCAST") %>%
  ggplot(aes(Measurement, fill = index)) + geom_bar() + 
  # EUCAST cut-off of 21mm is resistant
  geom_vline(aes(xintercept = 21, linetype = "resistant"), color = "black") +
  # CLSI cut-off of 25mm is susceptible
  geom_vline(aes(xintercept = 25, linetype = "susceptible"), color = "black") +
  labs(title = "Total sample count at various disk diffusion measurements", x = "Measurement (mm)", y = "Frequency", fill = "Dataset", 
       linetype = "Disk diffusion \nbreakpoints") +
  scale_linetype_manual(values = c("solid", "dotted"))
```

## UpSet plots for different lab typing methods to visualise resistance mechanism combinations
```{r}
# combine Flq_mutations into one column
cipro_res_mech <- cipro_antibiogram %>%
  pivot_longer(Flq_mutations, names_to = "Flq.resistance.mech", 
               values_to = "Resistance.mech.type") %>%
  separate(Resistance.mech.type, into = c("type_A", "type_B", "type_C"), sep = "[;]")

cipro_res_mech <- cipro_res_mech %>%
  pivot_longer(cols = names(cipro_res_mech)[26:28], values_to = "Resistance.mech.type") %>%
  drop_na(Resistance.mech.type) %>%
  select(-name) 
```

### UpSet plots 
#### Violin plots {.tabset .tabset-pills}
##### MIC (broth dilution) {.tabset .tabset-pills}
###### CLSI
```{r fig.height=7, fig.width=7}
cipro_res_mech %>%
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "CLSI") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard") +
  labs(title = "Fluoroquinolone resistance determinant combinations at various\nMIC (broth dilution) measurements")
```

###### EUCAST
```{r fig.height=7, fig.width=7}
cipro_res_mech %>%
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "EUCAST") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard") +
  labs(title = "Fluoroquinolone resistance determinant combinations at various\nMIC (broth dilution) measurements") 
```

##### MIC (agar dilution) 
###### EUCAST
```{r fig.width=7, fig.height=7}
# only EUCAST used as testing standard
cipro_res_mech %>%
  filter(Laboratory.Typing.Method == "MIC (agar dilution)") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard") +
  labs(title = "Fluoroquinolone resistance determinant combinations\nat various MIC (agar dilution) measurements") 
```

##### Disk diffusion {.tabset .tabset-pills}
###### CLSI
```{r fig.width=7, fig.height=7}
cipro_res_mech %>%
  filter(Laboratory.Typing.Method == "Disk diffusion") %>%
  filter(Testing.standard == "CLSI") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard") 
```

###### EUCAST
```{r fig.width=7, fig.height=7}
cipro_res_mech %>%
  filter(Laboratory.Typing.Method == "Disk diffusion") %>%
  filter(Testing.standard == "EUCAST") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard")
```

From the violin plots, one can see that some genomes with no mutations are resistant to ciprofloxacin. These genomes were filtered out and saved in a file called **cipro_res_no_mutations.tsv** and used in the subsequent step of extracting *gyrA* and *parC* gene sequences and checking for mutations. 

```{r}
# filter for resistant strains with no gyrA or parC mutations
cipro_res_no_mut <- cipro_res_mech %>%
  filter(Resistance.phenotype == "resistant") %>%
  filter(Resistance.mech.type == "-")

#write_delim(cipro_res_no_mut, file = "cipro_res_no_mutations.tsv", delim = "\t")
```

# Extraction of *gyrA* and *parC* gene sequences
Assembled *K. pneumoniae* genomes of the 26 datasets used in the analysis were retrieved from Massive M3 using the **copy_fasta.sh** job script and GyrA and ParC reference amino acid sequences were retrieved from the Kleborate GitHub repo (https://github.com/katholt/Kleborate/blob/master/kleborate/resBLAST.py). TBLASTN was used to find nucleotide sequences of those proteins - GenBank accession number **AJ292307.1** for *gyrA* and **MZ242286.1** for *parC*. Translate tool on Expasy (https://web.expasy.org/translate/) was used to trim the nucleotide sequences so that the amino acid sequence produced matched that from Kleborate.

The *gyrA* and *parC* reference nucleotide sequences were then combined into one FASTA file which was used as the query for BLASTN while the subjects were the genomes from the various datasets. This was run using **cipro_blast.sh** on Massive. The output contained all the blast hits for the two genes. To obtain the unique blast hits, **cipro_blast_unique_hits.sh** was run. All the unique hits from the different datasets were combined into one dataframe and a metadata file was used to convert run accession IDs into strain IDs. Afterwards, the unique hits dataframe was combined with the genomes + matched antibiogram dataframe (cipro_antibiogram) which will be used to filter out extracted *gyrA* and *parC* sequences. 

```{r}
# read unique hits data files for all ciprofloxacin datasets using the data.table package
# and bind them by rows into one dataframe
cipro_uniquehits <- list.files(path = "./cipro_blast_unique_hits", full.names = T, pattern = "*.tsv") 
cipro_uniquehits <- do.call(rbind, lapply(cipro_uniquehits, read.delim, sep="\t"))

# rename strain IDs to match genome IDs
cipro_uniquehits <- cipro_uniquehits %>%
  mutate(strain = str_replace(strain, "_NextSeq_500_paired_end_sequencing", "")) %>%
  mutate(strain = str_replace(strain, "_Illumina_MiSeq_paired_end_sequencing", "")) %>%
  mutate(strain = str_replace(strain, "#", "_")) %>%
  mutate(strain = str_replace(strain, "KGVETS-", "")) %>%
  mutate(strain = str_replace(strain, "KGVET-", "")) %>%
  mutate(strain = str_replace(strain, "Klebsiella_pneumoniae_", "")) %>%
  mutate(strain = str_replace(strain, "_Klebsiella_pneumoniae", "")) %>%
  mutate(strain = str_replace(strain, "Klebsiella_variicola_", "")) %>%
  mutate(strain = str_replace(strain, "Klebsiella_quasipneumoniae_", "")) %>%
  mutate(strain = str_replace(strain, "_final", "")) %>%
  mutate(strain = str_replace(strain, "_out", "")) %>%
  mutate(strain = str_replace(strain, "_S.*_L555", "")) %>%
  mutate(strain = case_when(grepl("^2019.*B$", strain) ~ str_sub(strain, 1, -2),
                            TRUE ~ strain))

# read and combine all metadata files
kleb_sample_metadata <- list.files(path = "./cipro_res_kleb/metadata/datasets/", full.names = T, pattern = "*.tsv") %>%
  lapply(read_delim) %>%
  bind_rows

# save kleb_sample_metadata to tsv file
#write_delim(kleb_sample_metadata, file = "./cipro_res_kleb/metadata/kleb_sample_metadata.tsv", delim = "\t")

# merge unique hits and metadata files to extract strain names
cipro_uniquehits <- cipro_uniquehits %>%
  left_join(kleb_sample_metadata, by = c("strain" = "run_accession")) %>%
  mutate(sample_alias = ifelse(is.na(sample_alias), strain, sample_alias)) %>%
  select(- strain) %>%
  relocate(sample_alias, .before = query.gene) 

#write_delim(cipro_uniquehits, file = "cipro_uniquehits.tsv", delim = "\t")

# merge cipro_uniquehits and cipro_antibiogram and remove strains with assembled genomes
# but no matched antibiogram data
cipro_antibiogram_uniquehits <- cipro_antibiogram %>%
  left_join(cipro_uniquehits, by = c("strain" = "sample_alias")) %>%
  relocate(query.gene:coverage, .after = Flq_mutations) %>%
  drop_na(species) 

# strains with antibiogram data but no genomes
cipro_antibiogram_no_genomes <- cipro_antibiogram_uniquehits %>%
  filter(is.na(contig))

#write_delim(cipro_antibiogram_no_genomes, file = "cipro_antibiogram_no_genomes.tsv", delim = "\t")

# remove strains with no genome sequences i.e. strains with antibiogram data but no genomes
cipro_antibiogram_uniquehits <- cipro_antibiogram_uniquehits %>%
  drop_na(contig)

# save cipro_antibiogram_uniquehits to tsv file
#write_delim(cipro_antibiogram_uniquehits, file = "cipro_antibiogram_uniquehits.tsv", delim = "\t")
```

*gyrA* and *parC* gene sequences were then extracted using **extract_blast_gyra.sh** and **extract_blast_parc.sh** job scripts respectively using the default cut-off for length (50%) and 95%, 90%, and finally 80% identity cut-off. **filter_gyrA_parC.py** was used to check whether *gyrA* and *parC* gene sequences were extracted for all genomes (fasta file headers were manually edited: .fasta contig -> .fasta_contig e.g. .fasta 6 -> .fasta_6") and rename the headers in the following pattern: ID_gene_contig. This was run on Massive using **filter_gyrA_parC.sh**. *gyrA* sequence of **KP_NORM_BLD_2011_75132** was added manually (after querying the genome against the reference using BLASTN) as it was not picked up by the job script.

Gaps were then removed from the resulting output, **filtered_gyrA_parC.fasta**, using `seqkit seq -g filtered_gyrA_parC.fasta > filtered_gyrA_parC_nogaps.fasta`and the presence of any duplicated IDs was checked using `seqkit rmdup -n < filtered_gyrA_parC_nogaps.fasta > filtered_gyrA_parC_nogaps_NR.fasta -D filtered_gyrA_parC_nogaps_NR_duplicate_info.txt -d filtered_gyrA_parC_nogaps_NR_duplicate_seqs.fasta`. *gyrA* and *parC* gene sequences were then separated using `seqkit grep -r -p gyrA filtered_gyrA_parC_nogaps_NR.fasta -o filtered_gyrA.fasta` and `seqkit grep -r -p parC filtered_gyrA_parC_nogaps_NR.fasta -o filtered_parC.fasta` respectively to obtain **filtered_gyrA.fasta** and **filtered_parC.fasta** as output files. Prodigal was then run to translate the *gyrA* and *parC* nucleotide sequences into amino acid sequences using the following commands respectively: `./prodigal.linux -i filtered_gyrA.fasta -o filtered_gyrA_gene.coords.gbk -a filtered_gyrA_aa.faa` and `./prodigal.linux -i filtered_parC.fasta -o filtered_parC_gene.coords.gbk -a filtered_parC_aa.faa`.

BLASTP was then run on Massive using **blast_gyra.sh** and **blast_parc.sh** whereby **filtered_gyrA_aa.faa** and **filtered_parC_aa.faa** were queried against **gyra_ref_aa_seq.faa** and **parc_ref_aa_seq.faa** to get blast outputs with the best hits, **gyrA_blast.tsv** and **parC_blast.tsv**. The amino acid sequences of the best hits were extracted using **extract_gyra_blast.py** and **extract_parc_blast.py** into **gyra_blast.faa** and **parc_blast.faa** respectively. The GyrA amino acid sequence of **KP_NORM_BLD_2011_75132** was added manually to the **gyra_blast.faa** file by first translating the nucleotide sequence using the Translate tool on Expasy (https://web.expasy.org/translate/) - 3'5' frame 1 was chosen. **gyra_blast.faa** and **parc_blast.faa** were then aligned to the reference sequences using mafft: `mafft parc_blast.faa/gyra_blast.faa > parc_blast_align.faa/gyra_blast_align.faa` to first align the sequences then `mafft --addfull parc_ref_aa_seq.faa/gyra_ref_aa_seq.faa parc_blast_align.faa/gyra_blast_align.faa > parc_blast_align_with_ref.faa/gyra_blast_align_with_ref.faa` to align them to the reference. The reference sequences were thereafter removed from the resulting alignment files. Afterwards, strains with WT GyrA/ParC, those with mutations in the proteins and with truncated amino acid sequences were summarised using **gyrA_mismatch_pos.py** and **parC_mismatch_pos.py**. The mutations - **gyrA_mutations.csv** and **parC_mutations.csv** - were then compared to the Kleborate output and mutations not in Kleborate but fall within the quinolone resistance-determining region (QRDR) i.e. position 67-106 for GyrA and 63 to 102 for ParC were saved in **mutations_not_in_kleborate.tsv**. Mutations different from those picked up by Kleborate were marked with a double asterisk (**).

```{r results='hide'}
# read GyrA mutations file
gyra_mutations <- read.csv("cipro_res_kleb/extracted_sequences/gyrA_mutations.csv", header = TRUE)

# select  sequence ID and mutations columns and edit sequence IDs to match strain IDs
gyra_mutations <- gyra_mutations %>%
  select(Sequence.ID, Mutations) %>%
  mutate(Sequence.ID = str_replace(Sequence.ID, "_gyrA_[0-9]+_[0-9]+$", "")) %>%
  mutate(Sequence.ID = str_replace(Sequence.ID, "_gyrA_NODE.*$", "")) %>%
  rename("strain" = "Sequence.ID", "gyra_mutations" = "Mutations")

# read ParC mutations file
parc_mutations <- read.csv("cipro_res_kleb/extracted_sequences/parC_mutations.csv", header = TRUE)

# select  sequence ID and mutations columns and edit sequence IDs to match strain IDs
parc_mutations <- parc_mutations %>%
  select(Sequence.ID, Mutations) %>%
  mutate(Sequence.ID = str_replace(Sequence.ID, "_parC_[0-9]+_[0-9]+$", "")) %>%
  mutate(Sequence.ID = str_replace(Sequence.ID, "_parC_chromosome_[0-9]+$", "")) %>%
  mutate(Sequence.ID = str_replace(Sequence.ID, "_parC_NODE.*$", "")) %>%
  rename("strain" = "Sequence.ID", "parc_mutations" = "Mutations")

# separate Flq_mutations into individual columns
kleborate_mutations <- cipro_antibiogram %>%
  select(strain, Flq_mutations) %>%
  separate(Flq_mutations, into = c("mut_A", "mut_B", "mut_C"), sep = "[;]") 

# create columns for GyrA and ParC mutations  
kleborate_mutations <- kleborate_mutations %>%
  mutate(gyra_mutations = case_when(grepl("^GyrA", mut_A) & is.na(mut_B) ~
                                      paste(mut_A),
                                    grepl("^GyrA", mut_A) & grepl("^GyrA", mut_B) ~
                                      paste(mut_A, mut_B, sep = ";"),
                                    grepl("^GyrA", mut_A) & grepl("^ParC", mut_B) ~
                                      paste(mut_A))) %>%
  mutate(parc_mutations = case_when(grepl("^ParC", mut_B) & is.na(mut_C) ~
                                      paste(mut_B),
                                    grepl("^GyrA", mut_A) & grepl("^GyrA", mut_B) & grepl("^ParC", mut_C) ~
                                      paste(mut_C))) %>%
  replace_na(list(gyra_mutations = "-", parc_mutations = "-"))

# compare kleborate mutations with gyra and parc mutations
# combine kleborate_mutations with gyra_mutations and parc_mutations
combined_mutations <- kleborate_mutations %>%
  left_join(gyra_mutations, by = "strain", suffix = c("_kleb", "_extract")) %>%
  left_join(parc_mutations, by = "strain", suffix = c("_kleb", "_extract")) %>%
  relocate(gyra_mutations_extract, .after = gyra_mutations_kleb) %>%
  replace_na(list(gyra_mutations_extract = "-", parc_mutations_extract = "-"))

# compare _kleb and _extract columns and bind logical output to combined_mutations
gyra_compare <- combined_mutations$gyra_mutations_kleb == combined_mutations$gyra_mutations_extract
parc_compare <- combined_mutations$parc_mutations_kleb == combined_mutations$parc_mutations_extract

combined_mutations <- combined_mutations %>%
  cbind(as.data.frame(gyra_compare), as.data.frame(parc_compare)) %>%
  relocate(gyra_compare, .after = gyra_mutations_extract)

# find which mutations are not in the Kleborate output
gyra_not_in_kleb <- combined_mutations %>%
  filter(gyra_compare == FALSE) %>%
  filter(gyra_mutations_extract != "-") %>%
  select(strain, gyra_mutations_kleb:gyra_compare)

parc_not_in_kleb <- combined_mutations %>%
  filter(parc_compare == FALSE) %>%
  filter(parc_mutations_extract != "-") %>%
  select(strain, parc_mutations_kleb:parc_compare)

# find mutations in the QRDR region i.e. between position 67 and 106 for GyrA and position 63 and
# 102 for ParC
# first for GyrA
gyra_in_qrdr <- gyra_not_in_kleb %>%
  select(strain, gyra_mutations_extract) %>%
  separate(gyra_mutations_extract, into = c("mut_A", "mut_B", "mut_C", "mut_D", "mut_E", "mut_F", "mut_G", "mut_H", "mut_I"), 
           sep = "[;]")

# number_range() allows one to set number range as regex
(gyra_qrdr_range <- number_range(67, 106))

gyra_in_qrdr <- gyra_in_qrdr %>%
  mutate(gyra_mutations = case_when(grepl(gyra_qrdr_range, mut_A) & is.na(mut_B) ~ paste(mut_A),
                                    grepl(gyra_qrdr_range, mut_A) & grepl(gyra_qrdr_range, mut_B) & is.na(mut_C) ~ 
                                      paste(mut_A, mut_B, sep = ";"),
                                    grepl(gyra_qrdr_range, mut_A) & grepl(gyra_qrdr_range, mut_B) & grepl(gyra_qrdr_range, mut_C) &
                                    grepl(gyra_qrdr_range, mut_D) ~ paste(mut_A, mut_B, mut_C, mut_D, mut_E, mut_F, mut_G, mut_H, mut_I, sep = ";"))) %>%
  replace_na(list(gyra_mutations = "-")) %>%
  select(strain, gyra_mutations) %>%
  filter(gyra_mutations != "-")

# now for ParC
parc_in_qrdr <- parc_not_in_kleb %>%
  select(strain, parc_mutations_extract) %>%
  separate(parc_mutations_extract, into = c("mut_A", "mut_B"), sep = "[;]")

(parc_qrdr_range <- number_range(63, 102))

parc_in_qrdr <- parc_in_qrdr %>%
  mutate(parc_mutations = case_when(grepl(parc_qrdr_range, mut_A) & is.na(mut_B) ~ paste(mut_A),
                                    grepl(parc_qrdr_range, mut_A) & grepl(parc_qrdr_range, mut_B) ~ 
                                      paste(mut_A, mut_B, sep = ";"))) %>%
  replace_na(list(parc_mutations = "-")) %>%
  select(strain, parc_mutations) %>%
  filter(parc_mutations != "-")

# then combine GyrA and ParC
mutations_not_in_kleborate <- gyra_in_qrdr %>%
  full_join(parc_in_qrdr, by = "strain") %>%
  mutate(Flq_mutations = case_when(grepl("^GyrA", gyra_mutations) & grepl("^ParC", parc_mutations) ~ 
                                     paste(gyra_mutations, parc_mutations, sep = ";"), 
                                   grepl("^GyrA", gyra_mutations) & is.na(parc_mutations) ~ 
                                     paste(gyra_mutations),
                                   grepl("^ParC", parc_mutations) & is.na(gyra_mutations) ~
                                     paste(parc_mutations))) %>%
  select(- c("gyra_mutations", "parc_mutations"))

#write_delim(mutations_not_in_kleborate, file = "mutations_not_in_kleborate.tsv", delim = "\t")
```

# Subsequent analysis of resistant strains and STs
Following data cleaning, preliminary analysis, and extraction of GyrA and ParC mutations, mutations in the two proteins that had not been included in the Kleborate output were added to cipro_antibiogram and the Flq_acquired column was also included in subsequent data analysis - some strains were resistant but had no mutations in GyrA or ParC therefore acquired fluoroquinolone resistance genes might be contributing to their resistant phenotype.

```{r}
# read antibiogram data again and filter for ciprofloxacin
antibiogram <- read.delim(file = "antibiogram_combined.tsv", 
                                header = TRUE, sep = "\t")

# edit: remove Omp mutations from analysis as based on violin + upset plots, they
# do not seem to contribute to ciprofloxacin resistance
cipro_antibiogram <- antibiogram %>%
  filter(Antibiotic == "ciprofloxacin") %>%
  select(strain:ST, Flq_mutations, Flq_acquired, Antibiotic:Measurement.Round) %>%
  mutate(Resistance.phenotype = case_when(grepl('MIC.*$', Laboratory.Typing.Method) & Measurement <= 0.25 ~ 
                                            "susceptible",
                                          grepl('MIC.*$', Laboratory.Typing.Method) & Testing.standard == 
                                            "EUCAST" & Measurement > 0.5 ~ "resistant",
                                          grepl('MIC.*$', Laboratory.Typing.Method) & Testing.standard == 
                                            "CLSI" & Measurement >= 1 ~ "resistant",
                                          Laboratory.Typing.Method == "Disk diffusion" & Testing.standard == 
                                            "EUCAST" & Measurement >= 25 ~ "susceptible",
                                          Laboratory.Typing.Method == "Disk diffusion" & Testing.standard == 
                                            "EUCAST" & Measurement < 22 ~ "resistant",
                                          Laboratory.Typing.Method == "Disk diffusion" & Testing.standard == 
                                            "CLSI" & Measurement >= 26 ~ "susceptible",
                                          Laboratory.Typing.Method == "Disk diffusion" & Testing.standard == 
                                            "CLSI" & Measurement <= 21 ~ "resistant",
                                          TRUE ~ "intermediate")) %>%
  # filter out strains with measurement of 0.12 and 0.75 (no such MIC measurements)
  filter(Measurement != 0.12 & Measurement != 0.75)

#write_delim(cipro_antibiogram, file = "cipro_antibiogram.tsv", delim = "\t")

# add mutations not in Kleborate output to cipro_antibiogram
# read mutations file
no_mut_kleborate <- read.delim(file = "mutations_not_in_kleborate.tsv",
                               header = TRUE, sep = "\t")

# combine antibiogram with mutations not in kleborate data file
cipro_antibiogram <- cipro_antibiogram %>%
  left_join(no_mut_kleborate, by = c("strain")) %>%
  mutate(Flq_mutations = case_when(Flq_mutations.y %in% NA ~ Flq_mutations.x, 
                                   Flq_mutations.x == "-" & Flq_mutations.y %in% NA ~ Flq_mutations.x,
                                   Flq_mutations.x == "-" ~ Flq_mutations.y,
                                   Flq_mutations.x != "-" & !is.na(Flq_mutations.y) & !Flq_mutations.x %in% Flq_mutations.y ~
                                     paste(Flq_mutations.x, Flq_mutations.y, sep = ";"),
                                   TRUE ~ Flq_mutations.x)) %>%
  select(- c(Flq_mutations.x, Flq_mutations.y)) %>%
  relocate(Flq_mutations, .after = ST)

# separate mutations and check for duplicates then combine them again into the Flq_mutations column
cipro_antibiogram <- cipro_antibiogram %>%
  pivot_longer(Flq_mutations, names_to = "Flq.resistance.mech", 
               values_to = "Resistance.mech.type") %>%
  separate(Resistance.mech.type, into = c("type_A", "type_B", "type_C", "type_D", "type_E", "type_F", "type_G", "type_H", "type_I"), 
           sep = "[;]")

cipro_antibiogram <- cipro_antibiogram %>%
  pivot_longer(cols = names(cipro_antibiogram)[27:35], values_to = "Resistance.mech.type") %>%
  drop_na(Resistance.mech.type) %>%
  select(-name) %>%
  filter(!duplicated(cbind(strain, Resistance.mech.type))) %>%
  pivot_wider(names_from = Flq.resistance.mech, values_from = Resistance.mech.type,
              values_fn = function(x) paste(x, collapse=";")) %>%
  relocate(Flq_mutations, .before = Flq_acquired)
```

## Summaries {.tabset .tabset-pills}
### Sample count per dataset
```{r}
cipro_antibiogram %>%
  group_by(index) %>%
  ggplot(aes(x = index)) +
  geom_bar(fill = "#ff8000", alpha = 0.6) +
  geom_text(aes(label = ..count..), stat = "count", nudge_y = 60) +
  labs(title = "Total sample count per dataset", y = "Total sample count", fill = "Lab typing method") +
  theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1))
```

### Total sample counts vs measurements for different lab typing methods {.tabset .tabset-pills}
#### MIC (broth dilution) {.tabset .tabset-pills}
##### CLSI
```{r}
cipro_antibiogram %>%
  group_by(Laboratory.Typing.Method) %>%
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "CLSI") %>% 
  ggplot(aes(x = factor(Measurement), fill = index)) + geom_bar() + 
  # CLSI cut-off of 1.0 mg/L is resistant
  geom_vline(aes(xintercept = 5, linetype = "resistant"), alpha = 0.6, color = "black") +
  # CLSI cut-off of 0.25 mg/L is susceptible
  geom_vline(aes(xintercept = 3, linetype = "susceptible"), color = "black") +
  labs(title = "Total sample count at various MIC (broth dilution) measurements", x = "Measurement (mg/L)", y = "Frequency", fill = "Dataset", 
       linetype = "MIC breakpoints") +
  scale_linetype_manual(values = c("solid", "dotted")) 
```

##### EUCAST
```{r fig.height=6}
cipro_antibiogram %>%
  group_by(Laboratory.Typing.Method) %>%
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "EUCAST") %>% 
  ggplot(aes(x = factor(Measurement), fill = index)) + geom_bar() + 
  # EUCAST cut-off of 1.0 mg/L is resistant
  geom_vline(aes(xintercept = 8, linetype = "resistant"), alpha = 0.6, color = "black") +
  # EUCAST cut-off of 0.25 mg/L is susceptible
  geom_vline(aes(xintercept = 5, linetype = "susceptible"), color = "black") +
  labs(title = "Total sample count at various MIC (broth dilution) measurements", x = "Measurement (mg/L)", y = "Frequency", fill = "Dataset", 
       linetype = "MIC breakpoints") +
  scale_linetype_manual(values = c("solid", "dotted")) 
```

#### MIC (agar dilution)
##### EUCAST
```{r}
# only EUCAST used as testing standard
cipro_antibiogram %>%
  group_by(Laboratory.Typing.Method) %>%
  filter(Laboratory.Typing.Method == "MIC (agar dilution)") %>% 
  ggplot(aes(x = factor(Measurement), fill = index)) + geom_bar() + 
  # EUCAST cut-off of 1.0 mg/L is resistant
  geom_vline(aes(xintercept = 6, linetype = "resistant"), alpha = 0.6, color = "black") +
  # EUCAST cut-off of 0.25 mg/L is susceptible
  geom_vline(aes(xintercept = 4, linetype = "susceptible"), color = "black") +
  labs(title = "Total sample count at various MIC (agar dilution) measurements", x = "Measurement (mg/L)", y = "Frequency", fill = "Dataset", 
       linetype = "MIC breakpoints") +
  # round off x-axis labels to 2dp
  scale_x_discrete(labels = function(x) round(as.numeric(x), digits = 2)) +
  scale_linetype_manual(values = c("solid", "dotted")) 
```

#### Disk diffusion {.tabset .tabset-pills}
##### CLSI
```{r}
cipro_antibiogram %>%
  group_by(Laboratory.Typing.Method) %>%
  filter(Laboratory.Typing.Method == "Disk diffusion") %>%
  filter(Testing.standard == "CLSI") %>%
  ggplot(aes(Measurement, fill = index)) + geom_bar() + 
  # CLSI cut-off of 21mm is resistant
  geom_vline(aes(xintercept = 21, linetype = "resistant"), color = "black") +
  # CLSI cut-off of 26mm is susceptible
  geom_vline(aes(xintercept = 26, linetype = "susceptible"), color = "black") +
  labs(title = "Total sample count at various disk diffusion measurements", x = "Measurement (mm)", y = "Frequency", fill = "Dataset", 
       linetype = "Disk diffusion \nbreakpoints") +
  scale_linetype_manual(values = c("solid", "dotted"))
```

##### EUCAST
```{r}
cipro_antibiogram %>%
  group_by(Laboratory.Typing.Method) %>%
  filter(Laboratory.Typing.Method == "Disk diffusion") %>%
  filter(Testing.standard == "EUCAST") %>%
  ggplot(aes(Measurement, fill = index)) + geom_bar() + 
  # EUCAST cut-off of 21mm is resistant
  geom_vline(aes(xintercept = 21, linetype = "resistant"), color = "black") +
  # CLSI cut-off of 25mm is susceptible
  geom_vline(aes(xintercept = 25, linetype = "susceptible"), color = "black") +
  labs(title = "Total sample count at various disk diffusion measurements", x = "Measurement (mm)", y = "Frequency", fill = "Dataset", 
       linetype = "Disk diffusion \nbreakpoints") +
  scale_linetype_manual(values = c("solid", "dotted"))
```

## UpSet plots for different lab typing methods to visualise resistance mechanism combinations
```{r}
# combine Flq_mutations and Flq_acquired into one column
# edit: remove Omp mutations from analysis as based on violin + upset plots, they
# do not seem to contribute to ciprofloxacin resistance
cipro_res_mech <- cipro_antibiogram %>%
  pivot_longer(Flq_mutations:Flq_acquired, names_to = "Flq.resistance.mech", 
               values_to = "Resistance.mech.type") %>%
  separate(Resistance.mech.type, into = c("type_A", "type_B", "type_C"), sep = "[;]")

cipro_res_mech <- cipro_res_mech %>%
  pivot_longer(cols = names(cipro_res_mech)[26:28], values_to = "Resistance.mech.type") %>%
  drop_na(Resistance.mech.type) %>%
  select(-name) 

# group gyrA, parC, and qnr genes
cipro_res_mech <- cipro_res_mech %>%
  group_res_mech(Resistance.mech.type)

# explicitly label strains with wildtype QRDR (gyrA/parC) and no acquired genes as such
cipro_res_mech <- cipro_res_mech %>%
  mutate(Resistance.mech.type = replace(Resistance.mech.type, Flq.resistance.mech=="Flq_mutations" & Resistance.mech.type=="-", "wt_QRDR")) %>%
  mutate(Resistance.mech.type = replace(Resistance.mech.type, Flq.resistance.mech=="Flq_acquired" & Resistance.mech.type=="-", "no_acquired_genes"))
```

### Summary bar plot of resistance mechanisms {.tabset .tabset-pills}
#### Count table
```{r}
# count number of occurrences of the different fluoroquinolone resistance determinants
summ_res_mech <- cipro_res_mech %>%
  filter(Resistance.mech.type != "wt_QRDR" & Resistance.mech.type != "no_acquired_genes") %>%
  group_by(Flq.resistance.mech, Resistance.mech.type) %>%
  summarise(count = n()) 

DT::datatable(summ_res_mech,
              width = "50%",
              extensions = "FixedHeader",
              rownames = FALSE,
              options = list(paging=TRUE, fixedHeader=TRUE, scrollX=TRUE))
```

#### Summary bar plot
```{r fig.width=9, fig.height=8}
summ_res_mech %>%
  ggplot(aes(Resistance.mech.type, count, fill = Flq.resistance.mech)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = count), stat = "identity", nudge_y = 60) +
  labs(title = "Total sample count per resistance determinant", x = "Resistance mechanism",
       y = "Total sample count", fill = "Fluoroquinolone resistance \nmechanism type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_fill_brewer(palette = "Accent")
```

### UpSet plots {.tabset .tabset-pills}
#### Violin plots {.tabset .tabset-pills}
##### MIC (broth dilution) {.tabset .tabset-pills}
###### CLSI
```{r fig.height=10, fig.width=12}
cipro_res_mech %>%
  filter(!(Resistance.mech.type == "no_acquired_genes")) %>% # exclude these rows as not informative
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "CLSI") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard") +
  labs(title = "Fluoroquinolone resistance determinant combinations at various MIC (broth dilution) measurements")
```

###### EUCAST
```{r fig.height=10, fig.width=13}
cipro_res_mech %>%
  filter(!(Resistance.mech.type == "no_acquired_genes")) %>% # exclude these rows as not informative
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "EUCAST") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard") +
  labs(title = "Fluoroquinolone resistance determinant combinations at various MIC (broth dilution) measurements") 
```

##### MIC (agar dilution) 
###### EUCAST
```{r fig.width=10, fig.height=9}
# only EUCAST used as testing standard
cipro_res_mech %>%
  filter(!(Resistance.mech.type == "no_acquired_genes")) %>% # exclude these rows as not informative
  filter(Laboratory.Typing.Method == "MIC (agar dilution)") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard") +
  labs(title = "Fluoroquinolone resistance determinant combinations\nat various MIC (agar dilution) measurements") 
```

##### Disk diffusion {.tabset .tabset-pills}
###### CLSI
```{r fig.width=8, fig.height=7}
cipro_res_mech %>%
  filter(!(Resistance.mech.type == "no_acquired_genes")) %>% # exclude these rows as not informative
  filter(Laboratory.Typing.Method == "Disk diffusion") %>%
  filter(Testing.standard == "CLSI") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard") 
```

###### EUCAST
```{r fig.width=11, fig.height=9}
cipro_res_mech %>%
  filter(!(Resistance.mech.type == "no_acquired_genes")) %>% # exclude these rows as not informative
  filter(Laboratory.Typing.Method == "Disk diffusion") %>%
  filter(Testing.standard == "EUCAST") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard")
```

#### Percent barcharts {.tabset .tabset-pills}
##### MIC (broth dilution) {.tabset .tabset-pills}
###### CLSI
```{r fig.height=10, fig.width=12}
cipro_res_mech %>%
  filter(!(Resistance.mech.type == "no_acquired_genes")) %>% # exclude these rows as not informative
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "CLSI") %>%
  upset_bar_plot("Resistance.mech.type") +
  labs(title = "Phenotypes for fluoroquinolone resistance determinant combinations")
```

###### EUCAST
```{r fig.height=10, fig.width=13}
cipro_res_mech %>%
  filter(!(Resistance.mech.type == "no_acquired_genes")) %>% # exclude these rows as not informative
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "EUCAST") %>%
  upset_bar_plot("Resistance.mech.type") +
  labs(title = "Phenotypes for fluoroquinolone resistance determinant combinations")
```

##### MIC (agar dilution) 
###### EUCAST
```{r fig.width=10, fig.height=8}
# only EUCAST used as testing standard
cipro_res_mech %>%
  filter(!(Resistance.mech.type == "no_acquired_genes")) %>% # exclude these rows as not informative
  filter(Laboratory.Typing.Method == "MIC (agar dilution)") %>%
  upset_bar_plot("Resistance.mech.type") +
  labs(title = "Phenotypes for fluoroquinolone resistance determinant combinations")
```

##### Disk diffusion {.tabset .tabset-pills}
###### CLSI
```{r fig.width=7, fig.height=6}
cipro_res_mech %>%
  filter(!(Resistance.mech.type == "no_acquired_genes")) %>% # exclude these rows as not informative
  filter(Laboratory.Typing.Method == "Disk diffusion") %>%
  filter(Testing.standard == "CLSI") %>%
  upset_bar_plot("Resistance.mech.type") +
  labs(title = "Phenotypes for fluoroquinolone resistance\ndeterminant combinations")
```

###### EUCAST
```{r fig.width=9, fig.height=8}
cipro_res_mech %>%
  filter(!(Resistance.mech.type == "no_acquired_genes")) %>% # exclude these rows as not informative
  filter(Laboratory.Typing.Method == "Disk diffusion") %>%
  filter(Testing.standard == "EUCAST") %>%
  upset_bar_plot("Resistance.mech.type") +
  labs(title = "Phenotypes for fluoroquinolone resistance\ndeterminant combinations")
```

## ST analysis
### Resistant strains displaying fluoroquinolone resistance mechanisms {.tabset .tabset-pills}
#### ST count 
```{r}
# count STs with resistant phenotype
top_35_st_res <- cipro_antibiogram %>%
  filter(Resistance.phenotype == "resistant") %>%
  filter(Flq_mutations != "-" | Flq_acquired != "-") %>%
  group_by(ST) %>%
  summarise(sample_count = n()) %>%
  as.data.frame() %>%
  arrange(desc(sample_count)) %>%
  head(35) %>%
  mutate(ST = factor(ST, levels = ST, ordered = TRUE))

DT::datatable(top_35_st_res, 
              width = "50%",
              extensions = "FixedHeader",
              rownames = FALSE,
              options = list(paging=TRUE, fixedHeader=TRUE, scrollX=TRUE))
```

#### Summary bar plot
```{r fig.width=10, fig.height=8}
# Blues palette with 9 colors
colour_palette <- brewer.pal(9, "Blues") 

# Add more colors to this palette :
colour_palette <- colorRampPalette(colour_palette)(35)

top_35_st_res %>%
  ggplot(aes(y = ST, x = sample_count, fill = ST, label = sample_count)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(hjust = -0.5) +
  scale_y_discrete(limits = rev) +
  ggtitle("Top STs with fluoroquinolone resistance") +
  labs(x = "Sample count") +
  # reverse order of colour palette
  scale_fill_manual(values = rev(colour_palette)) +
  theme_minimal() +
  theme(legend.position = "none",
  axis.text = element_text(size = 10)) +
  coord_cartesian(clip = "off")
```

#### Resistant STs associated with specific fluoroquinolone resistance mechanisms {.tabset .tabset-pills}
##### Proportion table
```{r}
st_totals <- cipro_antibiogram %>%
  group_by(ST) %>%
  summarise(N=n())

st_res_mech <- cipro_res_mech %>%
  filter(Resistance.phenotype == "resistant") %>%
  filter(Resistance.mech.type != "wt_QRDR", Resistance.mech.type != "no_acquired_genes") %>%
  count(ST, Resistance.mech.type) %>% 
  left_join(st_totals) %>% # ST total counts
  mutate(prop = n/N) %>% # proportion per ST
  mutate(se = sqrt(prop*(1-prop)/N)) %>%
  mutate(p_lowerCI = prop-1.96*se, p_upperCI = prop+1.96*se) %>%
  mutate(p_lowerCI = replace(p_lowerCI, p_lowerCI<0, 0)) %>%
  mutate(p_upperCI = replace(p_upperCI, p_upperCI>1, 1)) %>%
  select(-se) %>%
  mutate(ST = factor(ST, levels = unique(ST)))

DT::datatable(st_res_mech, 
              extensions = "FixedHeader",
              rownames = FALSE,
              options = list(paging=TRUE, fixedHeader=TRUE, scrollX=TRUE))
```

##### Heatmap of the proportions of resistance determinants per ST
```{r}
st_res_mech %>%
  filter(ST %in% st_res_mech$ST[st_res_mech$prop>0.1 & st_res_mech$N>50]) %>%
  ggplot(aes(x=ST, y = Resistance.mech.type)) +
  geom_tile(aes(fill=prop)) +
  labs(subtitle = "Fluoroquinolone resistance mechanisms in top resistant STs", 
       y = "Fluoroquinolone resistance mechanisms", fill = "Proportion per \nST") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8, colour = "black")) + 
  scale_fill_gradient(low = "white", high = "red")
```

##### Heatmap/UpSet plot showing %proportion of resistance determinant combinations per ST {.tabset .tabset-pills}
###### Proportion table
```{r}
# separate resistance determinants
cipro_res_mech_prop <- cipro_antibiogram %>%
  separate(Flq_mutations, into = c("mut_A", "mut_B", "mut_C"), sep = "[;]") %>%
  separate(Flq_acquired, into = c("acq_A", "acq_B", "acq_C"), sep = "[;]") 

# group determinants based on statistical analysis
cipro_res_mech_prop <- cipro_res_mech_prop %>%
  group_res_mech(mut_A) %>%
  group_res_mech(mut_B) %>%
  group_res_mech(mut_C) %>%
  group_res_mech(acq_A) %>%
  group_res_mech(acq_B) %>%
  group_res_mech(acq_C) 

# combine previously separated resistance determinants columns
cipro_res_mech_prop <- cipro_res_mech_prop %>%
  unite("Flq_mutations", mut_A, mut_B, mut_C, sep = ";", na.rm = TRUE) %>%
  unite("Flq_acquired", acq_A, acq_B, acq_C, sep = ";", na.rm = TRUE) 

# combine all the resistance determinants columns using a ";" excluding rows without
# mutations 
cipro_res_mech_prop <- cipro_res_mech_prop %>%
  # replace "-" with NA
  replace_with_na(replace = list(Flq_mutations = "-", Flq_acquired = "-")) %>%
  unite("res_mech", Flq_mutations:Flq_acquired, sep = ";", remove = FALSE, na.rm = TRUE) %>%
  # replace NA with "-"
  replace_na(list(Flq_mutations = "-", Flq_acquired = "-")) 

# replace empty cells with "-"
cipro_res_mech_prop$res_mech <- sub("^$", "-", cipro_res_mech_prop$res_mech)

# calculate proportions of different combinations
st_res_mech_prop <- cipro_res_mech_prop %>%
  filter(Resistance.phenotype == "resistant") %>%
  filter(res_mech != "-") %>%
  group_by(ST) %>%
  count(ST, res_mech) %>%
  left_join(st_totals) %>% # ST total counts
  mutate(prop = n/N) %>% # proportion per ST
  mutate(se = sqrt(prop*(1-prop)/N)) %>%
  mutate(p_lowerCI = prop-1.96*se, p_upperCI = prop+1.96*se) %>%
  mutate(p_lowerCI = replace(p_lowerCI, p_lowerCI<0, 0)) %>%
  mutate(p_upperCI = replace(p_upperCI, p_upperCI>1, 1)) %>%
  select(-se) %>%
  mutate(ST = factor(ST, levels = unique(ST))) %>%
  arrange(res_mech) 

DT::datatable(st_res_mech_prop, 
              extensions = "FixedHeader",
              rownames = FALSE,
              options = list(paging=TRUE, fixedHeader=TRUE, scrollX=TRUE))
```

###### Heatmap of different combinations per ST
```{r fig.width=11, fig.height=9}
# create heatmap/upset plot of different combinations per ST
st_res_mech_prop %>%
  filter(ST %in% st_res_mech$ST[st_res_mech$prop>0.1 & st_res_mech$N>50]) %>%
  summarise(ST, res_mech, prop) %>%
  ggplot(aes(x=res_mech, y=ST, fill=prop)) +
  geom_tile() +
  axis_combmatrix(sep = ";") +
  theme_combmatrix(combmatrix.panel.line.size = 0.8) + 
  scale_fill_gradient(low = "white", high = "red") +
  labs(x = "Resistance determinants", fill = "Proportion per\nST",
       title = "Proportions of fluoroquinolone resistance determinant combinations per ST")
```

### Resistant strains displaying no fluoroquinolone resistance mechanisms {.tabset .tabset-pills}
#### Count table
```{r}
no_res_mech <- cipro_antibiogram %>%
  filter(Flq_mutations == "-", Flq_acquired == "-") %>%
  filter(Resistance.phenotype == "resistant")

st_no_res_mech <- no_res_mech %>%
  group_by(ST) %>%
  summarise(sample_count = n()) %>%
  as.data.frame() %>%
  arrange(desc(sample_count)) %>%
  mutate(ST = factor(ST, levels = ST, ordered = TRUE))

DT::datatable(st_no_res_mech,
              width = "50%",
              extensions = "FixedHeader",
              rownames = FALSE,
              options = list(paging=TRUE, fixedHeader=TRUE, scrollX=TRUE))
```

#### Bar plot
```{r}
st_no_res_mech %>%
  filter(sample_count > 2) %>%
  ggplot(aes(ST, sample_count)) +
  geom_bar(stat = "identity", fill = "blue", alpha = 0.7) +
  labs(title = "Resistant STs with no fluoroquinolone resistance mechanisms", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Susceptible strains displaying fluoroquinolone resistance mechanisms {.tabset .tabset-pills}
#### Count table
```{r}
sus_res_mech <- cipro_res_mech %>%
  filter(Resistance.phenotype == "susceptible") %>%
  filter(Resistance.mech.type != "wt_QRDR", Resistance.mech.type != "no_acquired_genes")

st_sus_res_mech <- sus_res_mech %>%
  group_by(ST) %>%
  summarise(res_mech = Resistance.mech.type) %>%
  as.data.frame() %>%
  count(ST, res_mech, sort = TRUE) %>%
  mutate(ST = factor(ST, levels = unique(ST)))

DT::datatable(st_sus_res_mech,
              width = "50%",
              extensions = "FixedHeader",
              rownames = FALSE,
              options = list(paging=TRUE, fixedHeader=TRUE, scrollX=TRUE))
```

#### Bar plot
```{r}
st_sus_res_mech %>%
  filter(n > 1) %>%
  ggplot(aes(ST, n, fill = res_mech)) +
  geom_bar(stat = "identity") +
  labs(title = "Fluoroquinolone resistance mechanisms employed by the top susceptible STs", y = "Count", fill = "Fluoroquinolone resistance\nmechanisms") +
  # put legend in one column
  guides(fill = guide_legend(ncol = 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### UpSet plots {.tabset .tabset-pills}
##### MIC (broth dilution) {.tabset .tabset-pills}
###### CLSI
```{r fig.width=7, fig.height=6}
sus_res_mech %>%
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "CLSI") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard") +
  labs(title = "Fluoroquinolone resistance determinant combinations\nat various MIC (broth dilution) measurements") 
```

###### EUCAST
```{r fig.width=7, fig.height=6}
sus_res_mech %>%
  filter(Laboratory.Typing.Method == "MIC (broth dilution)") %>%
  filter(Testing.standard == "EUCAST") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard") +
  labs(title = "Fluoroquinolone resistance determinant combinations\nat various MIC (broth dilution) measurements") 
```

##### MIC (agar dilution)
###### EUCAST
```{r fig.height=6}
# only EUCAST used as testing standard
sus_res_mech %>%
  filter(Laboratory.Typing.Method == "MIC (agar dilution)") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard") +
  labs(title = "Fluoroquinolone resistance determinant combinations\nat various MIC (agar dilution) measurements") 
```

##### Disk diffusion 
###### EUCAST
```{r fig.height=6}
# only EUCAST has values
sus_res_mech %>%
  filter(Laboratory.Typing.Method == "Disk diffusion") %>%
  filter(Testing.standard == "EUCAST") %>%
  upset_violin_plot("Resistance.mech.type", "Laboratory.Typing.Method", "Testing.standard")
```

```{r proportion, eval=FALSE, include=FALSE}
# statistical analysis
# proportion test

# calculate frequencies of different resistance mechanisms for the different phenotypes
# including numbers for the wildtype groups for comparison to compare resistance frequency 
# amongst strains with the determinant vs strains without
mech_freq <- cipro_res_mech %>%
  group_by(Resistance.mech.type, Resistance.phenotype) %>%
  filter(Resistance.phenotype != "intermediate") %>%
  summarise(n = n()) %>%
  ungroup() %>%
  group_by(Resistance.mech.type) %>%
  pivot_wider(names_from = Resistance.phenotype, values_from = n) 

# convert na values to 0
mech_freq[is.na(mech_freq)] <- 0

# find totals of different resistance mechanisms
mech_freq <- mech_freq %>%
  mutate(total = sum(c(resistant, susceptible)))


# compare each gyrA/parC SNP vs wildtype QRDR
qrdr_wt <- mech_freq %>% filter(Resistance.mech.type == "wt_QRDR")
qrdr_mut <- mech_freq %>% 
  filter(startsWith(Resistance.mech.type, "GyrA") | startsWith(Resistance.mech.type, "ParC")) %>%
  mutate(wt_res = qrdr_wt[2], wt_total=qrdr_wt[4]) # add columns with the wildtype counts for res & total
  
# run prop.test() to each row, comparing the resistant counts (x) with the total counts (n)
qrdr_prop_out <- dapply(as.matrix(qrdr_mut), MARGIN = 1, 
       FUN = function(y) broom::tidy(prop.test(x=c(as.numeric(y[2]), as.numeric(y[5])), n=c(as.numeric(y[4]), as.numeric(y[6])))))

# combine qrdr_mut and qrdr_prop_out
qrdr_mut_freq_prop <- cbind(qrdr_mut, qrdr_prop_out) %>%
  select(resistant, total, estimate1, estimate2, p.value) %>%
  mutate(p.value=as.numeric(p.value))

qrdr_mut_freq_prop

# specific mutations observed more than once, significantly associated with elevated resistance (vs wildtype QRDR)
view(qrdr_mut_freq_prop %>% filter(total>1) %>% filter(p.value < 0.05))

# the data provides evidence that ANY mutation at the 4 codons we track is associated with resistance, even though for some individual mutations we only have one observation:
# check gyrA83, there are 5 mutations observed at this codon (A, F, I, L, Y)
# 4 have large numbers (10) and significant p-values, gyrA-83A is observed only once but is also resistant
qrdr_mut_freq_prop %>% filter(startsWith(Resistance.mech.type,"GyrA-83"))

# Similar for GyrA-87, except that GyrA87-H is only seen in one strain and it is sensitive, so exclude this one (not a borderline phenotype)
qrdr_mut_freq_prop %>% filter(startsWith(Resistance.mech.type,"GyrA-87"))

# ParC-80 and ParC-84 look the same as for GyrA-83, ie all mutations at the site have same effect can be grouped together
qrdr_mut_freq_prop %>% filter(startsWith(Resistance.mech.type,"ParC-80"))
qrdr_mut_freq_prop %>% filter(startsWith(Resistance.mech.type,"ParC-84"))

# based on the statistical analysis, QRDR mutations will be updated to have just 4 types 
# (GyrA-83, GyrA-87, ParC-80, ParC-84) instead of 19 most of which are rare


# compare each Omp mutations vs wildtype Omp
#omp_wt <- mech_freq %>% filter(Resistance.mech.type == "wt_Omp")
#omp_mut <- mech_freq %>% 
  #filter(startsWith(Resistance.mech.type, "Omp")) %>%
  #mutate(wt_res = omp_wt[2], wt_total=omp_wt[4]) # add columns with the wildtype counts for res & total

# run prop.test() to each row, comparing the resistant counts (x) with the total counts (n)
#omp_prop_out <- dapply(as.matrix(omp_mut), MARGIN = 1, 
       #FUN = function(y) broom::tidy(prop.test(x=c(as.numeric(y[2]), as.numeric(y[5])), n=c(as.numeric(y[4]), as.numeric(y[6])))))

#omp_mut_freq_prop <- cbind(omp_mut, omp_prop_out) %>% 
  #select(resistant, total, estimate1, estimate2, p.value) %>%
  #mutate(p.value=as.numeric(p.value))

# all four mutations (truncation of either ompK35 or ompK36; and insertion of GD or TD in OmpK36, are all significantly associated with resistance)
#view(omp_mut_freq_prop)


# compare acquired vs no acquired genes
no_acquired <- mech_freq %>% filter(Resistance.mech.type == "no_acquired_genes")
acquired_genes <- mech_freq %>% 
  filter(startsWith(Resistance.mech.type, "qnr") | startsWith(Resistance.mech.type, "qep") ) %>%
  mutate(wt_res = no_acquired[2], wt_total=no_acquired[4]) # add columns with the wildtype counts for res & total

# run prop.test() to each row, comparing the resistant counts (x) with the total counts (n)
acquired_genes_prop_out <- dapply(as.matrix(acquired_genes), MARGIN = 1, 
       FUN = function(y) broom::tidy(prop.test(x=c(as.numeric(y[2]), as.numeric(y[5])), n=c(as.numeric(y[4]), as.numeric(y[6])))))

acquired_genes_freq_prop <- cbind(acquired_genes, acquired_genes_prop_out) %>% 
  select(resistant, total, estimate1, estimate2, p.value) %>%
  mutate(p.value=as.numeric(p.value))

view(acquired_genes_freq_prop)

# from the statistical analysis, the following can be combined:
# qepA2 (n=1, which is resistant) with qepA2* (n=17, all of which are resistant)
# qnrA1 and qnrA1^ (note the ^ means they are identical at protein level anyway) as all are resistant
# qnrB1.v1, qnrB1.v2, qnrB1.v2^ for similar reasons
# qnrB19, qnrB19^
# qnrB2.1, qnrB2.2
# qnrS1, qnrS1?, qnrS1*
# could also combine all the 'rare' genes that are only seen 1-2 times each but are always resistant:
# -> qnrB7, qnrE2, qnrS2, qnrVC1, qnrVC6
```

# Appendix
```{r get-labels, echo = FALSE}
labs = knitr::all_labels()
labs = labs[labs %in% c("functions", "proportion")]
```
```{r all-code, ref.label=labs, eval=FALSE}
```

# Session information
```{r}
sessionInfo()
```
